---
title: "EDA on mtcars"
author: "JAYANAGARKAR JANHAVI PRAKASH"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
---

```{r}
data(mtcars)
View(mtcars)
head(mtcars)
# Summary statistics
summary(mtcars)
# Structure of the dataset
str(mtcars)
```

```{r}
# Pairwise scatterplot matrix
pairs(mtcars)
```

```{r}
# Boxplot of mpg grouped by number of cylinders
boxplot(mpg ~ cyl, data = mtcars, col = "lightblue", main = "Boxplot of MPG by Number of Cylinders")
```

```{r}
# Scatterplot of mpg vs. horsepower
plot(mtcars$hp, mtcars$mpg, main = "Scatterplot of MPG vs. Horsepower", xlab = "Horsepower", ylab = "MPG")
# Correlation matrix
cor(mtcars)
```

```{r}
# Load necessary libraries
library(ggplot2)
# Histogram of mpg
hist(mtcars$mpg, col = "skyblue", main = "Histogram of MPG", xlab = "MPG")
```

```{r}
# Density plot of mpg
plot(density(mtcars$mpg), main = "Density Plot of MPG", xlab = "MPG")
```

```{r}
# Bar plot of cylinders count
barplot(table(mtcars$cyl), col = "lightgreen", main = "Bar Plot of Cylinders Count", xlab = "Number of Cylinders", ylab = "Count")
```

```{r}
# Pie chart of transmission types
pie(table(mtcars$am), col = c("skyblue", "lightcoral"), main = "Pie Chart of Transmission Types")
library(ggplot2)
```

```{r}
# ggplot2: Scatterplot of mpg vs. horsepower with color by cylinders
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point() +
  labs(title = "Scatterplot of MPG vs. Horsepower (Color by Cylinders)", x = "Horsepower", y = "MPG")
```

```{r}
# ggplot2: Boxplot of mpg grouped by number of cylinders
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot() +
  labs(title = "Boxplot of MPG by Number of Cylinders", x = "Number of Cylinders", y = "MPG")
```

```{r}
# ggplot2: Histogram of mpg with density plot
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density(color = "red") +
  labs(title = "Histogram with Density Plot of MPG", x = "MPG")
```

```{r}
# ggplot2: Bar plot of cylinders count
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Bar Plot of Cylinders Count", x = "Number of Cylinders", y = "Count")


```

```{r}
# ggplot2: Pie chart of transmission types
ggplot(mtcars, aes(x = "", fill = factor(am))) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Pie Chart of Transmission Types")
```

```{r}
# ggplot2: Faceted scatterplot matrix
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_grid(. ~ cyl) +
  labs(title = "Faceted Scatterplot Matrix of MPG vs. Horsepower by Cylinders", x = "MPG", y = "Horsepower")
```

```{r}
# Violin plot of mpg by number of cylinders
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_violin() +
  labs(title = "Violin Plot of MPG by Number of Cylinders", x = "Number of Cylinders", y = "MPG")
```

```{r}
# Correlation heatmap
library(corrplot)
corr_matrix <- cor(mtcars)
corr_matrix
corrplot(corr_matrix, method = "color")
```

```{r}
# Scatterplot matrix with ggplot2
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_grid(. ~ cyl) +
  labs(title = "Scatterplot Matrix of MPG vs. Horsepower by Cylinders", x = "MPG", y = "Horsepower")
```

```{r}
# 3D Scatterplot of mpg, hp, and wt
library(scatterplot3d)
scatterplot3d(mtcars$mpg, mtcars$hp, mtcars$wt, color = "blue", main = "3D Scatterplot")
```

```{r}
# Parallel coordinates plot
library(GGally)
ggpairs(mtcars, columns = c("mpg", "hp", "wt", "cyl"), title = "Parallel Coordinates Plot")
```

```{r}
# Interactive scatterplot with plotly
library(plotly)
plot_ly(data = mtcars, x = ~hp, y = ~mpg, color = ~cyl, type = "scatter", mode = "markers") %>%
  layout(title = "Interactive Scatterplot of MPG vs. Horsepower", xaxis = list(title = "Horsepower"), yaxis = list(title = "MPG"))
```

```{r}
#MULTIVARIATE ANALYSIS
# Load necessary libraries
library(ggplot2)
library(FactoMineR)
library(factoextra)

# Load the mtcars dataset
data(mtcars)
```

```{r}
# Select relevant variables for PCA
mtcars_pca_data <- mtcars[, c("mpg", "hp", "wt", "qsec")]
mtcars_pca_data
```
```{r}
# Standardize the data for PCA
mtcars_pca_standardized <- scale(mtcars_pca_data)
mtcars_pca_standardized
```

```{r}
# Perform PCA
mtcars_pca <- PCA(mtcars_pca_standardized, graph = FALSE)
mtcars_pca 
```

```{r}
# Scree plot to identify the number of principal components
fviz_screeplot(mtcars_pca, addlabels = TRUE, ylim = c(0, 40))
```



```{r}
# Variables factor map (scatterplot of variables in the space of the first two principal components)
fviz_pca_var(mtcars_pca, col.var = "contrib", gradient.cols = c("blue", "red"), title = "PCA Variables Factor Map")
```
```{r}
#REGRESSION ANALYSIS
# Fit a multiple linear regression model
model <- lm(mpg ~ hp + wt + qsec, data = mtcars)
model
```

```{r}
# Summary of the regression model
summary(model)
```
Conclusion:The model's R-squared value indicates that approximately 83% of the variability in miles per gallon can be explained by the predictors included in the model.


```{r}
# Diagnostic plots for the regression model
par(mfrow = c(2, 2))
plot(model)
```

```{r}
# Residuals vs. Fitted plot
plot(model$fitted.values, model$residuals, main = "Residuals vs. Fitted",
     xlab = "Fitted values", ylab = "Residuals", col = "blue", pch = 20)
```

```{r}
# Q-Q plot of residuals
qqnorm(model$residuals, main = "Normal Q-Q Plot of Residuals")
qqline(model$residuals)
```

```{r}
# Scale-Location plot (Square root of absolute residuals vs. Fitted values)
sqrt_abs_residuals <- sqrt(abs(model$residuals))
plot(model$fitted.values, sqrt_abs_residuals, main = "Scale-Location Plot",
     xlab = "Fitted values", ylab = "Square root of absolute residuals", col = "blue", pch = 20)
```

```{r}
# Residuals vs. Leverage plot
plot(model, which = 4)
```
```{R}
#OUTLIERS DETECTION

variables_of_interest <- c("mpg", "hp", "wt", "qsec")  # Define variables of interest for outlier detection

# Boxplot for outlier detection
boxplot(mtcars[, variables_of_interest], main = "Boxplot of Variables for Outlier Detection")

# Outlier identification using boxplot
outliers <- boxplot.stats(mtcars$mpg)$out
outliers
```
```{R}
#CLUSTER ANALYSIS
# Define variables for clustering
variables_for_clustering <- c("mpg", "hp", "wt")
variables_for_clustering
```

```{r}
# Standardize the data
scaled_data <- scale(mtcars[, variables_for_clustering])
scaled_data 
```

```{r}
# Perform K-means clustering
set.seed(123)  # For reproducibility
kmeans_result <- kmeans(scaled_data, centers = 3)
kmeans_result 
```

```{r}
# Add cluster assignments to the dataset
mtcars$cluster <- as.factor(kmeans_result$cluster)
mtcars$cluster 
```

```{r}
# Visualize clusters
ggplot(mtcars, aes(x = hp, y = mpg, color = cluster)) +
  geom_point() +
  labs(title = "K-means Clustering of Cars (mpg vs. hp)", x = "Horsepower", y = "Miles Per Gallon") +
  theme_minimal() +
  theme(legend.position = "top")
```
`

```{r}
#MACHINE LEARNING MODEL
# Load necessary libraries
library(caret)
```

```{r}
# Define features (predictors) and target variable
X <- mtcars[, c("hp", "wt", "qsec")]  # Features: horsepower, weight, quarter-mile time
y <- mtcars$mpg  # Target variable: miles per gallon
```

```{r}
# Split data into training and testing sets
set.seed(123)  # For reproducibility
train_indices <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_indices, ]
y_train <- y[train_indices]
X_test <- X[-train_indices, ]
y_test <- y[-train_indices]
```

```{r}
# Train a linear regression model
lm_model <- lm(y_train ~ ., data = data.frame(X_train))
lm_model 
```

```{r}
# Make predictions on the test set
predictions <- predict(lm_model, newdata = data.frame(X_test))
predictions
```
conclusion:
Mazda RX4 Wag: The predicted miles per gallon for the Mazda RX4 Wag is approximately 21.88 mpg. This means that according to the regression model, the Mazda RX4 Wag is estimated to achieve around 21.88 miles per gallon in terms of fuel efficiency.

Duster 360: The predicted miles per gallon for the Duster 360 is approximately 16.11 mpg. This suggests that the Duster 360 is estimated to have a lower fuel efficiency compared to the Mazda RX4 Wag, with around 16.11 miles per gallon.

Toyota Corolla: The predicted miles per gallon for the Toyota Corolla is approximately 27.81 mpg. This indicates that the Toyota Corolla is estimated to be more fuel-efficient compared to both the Mazda RX4 Wag and the Duster 360, achieving around 27.81 miles per gallon.

Ford Pantera L: The predicted miles per gallon for the Ford Pantera L is approximately 16.99 mpg. Similar to the Duster 360, the Ford Pantera L is estimated to have lower fuel efficiency compared to the Mazda RX4 Wag and the Toyota Corolla, with around 16.99 miles per gallon.

```{r}
# Evaluate model performance
accuracy <- sqrt(mean((predictions - y_test)^2))
accuracy
```
conclusion:
An RMSE of 3.26 suggests that the model's predictions have a moderate level of error in estimating the miles per gallon for the cars in the mtcars dataset.

```{r}
# Install and load necessary libraries
install.packages("shinydashboard")
library(shiny)
library(shinydashboard)
library(ggplot2)
library(plotly)
library(caret)
library(cluster)
```

```{r}
# Load the mtcars dataset
data(mtcars)

# Define UI components
ui <- dashboardPage(
  dashboardHeader(title = "Interactive EDA Dashboard"),
  dashboardSidebar(
    sidebarMenu(
      menuItem("Scatterplot", tabName = "scatterplot", icon = icon("chart-scatter")),
      menuItem("Boxplot", tabName = "boxplot", icon = icon("box")),
      menuItem("Analysis", tabName = "analysis", icon = icon("bar-chart-o"))
    )
  ),
  dashboardBody(
    tabItems(
      tabItem(tabName = "scatterplot",
              fluidRow(
                box(title = "Scatterplot Options",
                    selectInput("x_var", "X-axis variable:", choices = c("hp", "wt", "qsec")),
                    selectInput("y_var", "Y-axis variable:", choices = c("mpg", "hp", "wt", "qsec")),
                    actionButton("update_scatterplot", "Update Scatterplot", icon("refresh"))
                )
              ),
              fluidRow(
                plotlyOutput("scatterplot_output")
              )
      ),
      tabItem(tabName = "boxplot",
              fluidRow(
                box(title = "Boxplot Options",
                    selectInput("box_y_var", "Y-axis variable:", choices = c("mpg", "hp", "wt", "qsec")),
                    actionButton("update_boxplot", "Update Boxplot", icon("refresh"))
                )
              ),
              fluidRow(
                plotlyOutput("boxplot_output")
              )
      ),
      tabItem(tabName = "analysis",
              fluidRow(
                box(title = "Analysis Options",
                    radioButtons("analysis_type", "Analysis Type:",
                                 choices = c("Linear Regression", "K-means Clustering")),
                    conditionalPanel(
                      condition = "input.analysis_type == 'K-means Clustering'",
                      numericInput("num_clusters", "Number of Clusters:", value = 3, min = 1, max = 10)
                    ),
                    actionButton("run_analysis", "Run Analysis", icon("play-circle-o"))
                )
              ),
              fluidRow(
                verbatimTextOutput("analysis_results")
              )
      )
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  analysis_data <- reactiveVal(NULL)
  
  observeEvent(input$update_scatterplot, {
    output$scatterplot_output <- renderPlotly({
      gg <- ggplot(mtcars, aes_string(x = input$x_var, y = input$y_var)) +
        geom_point() +
        labs(title = paste("Scatterplot of", input$x_var, "vs.", input$y_var))
      ggplotly(gg)
    })
  })
  
  observeEvent(input$update_boxplot, {
    output$boxplot_output <- renderPlotly({
      gg <- ggplot(mtcars, aes_string(x = 1, y = input$box_y_var)) +
        geom_boxplot() +
        labs(title = paste("Boxplot of", input$box_y_var))
      ggplotly(gg)
    })
  })
  
  observeEvent(input$run_analysis, {
    if (input$analysis_type == "Linear Regression") {
      model <- train(as.formula(paste(input$y_var, "~ .")), data = mtcars, method = "lm")
      analysis_data(summary(model)$coef)
    } else if (input$analysis_type == "K-means Clustering") {
      clusters <- kmeans(mtcars[, c(input$x_var, input$y_var)], centers = input$num_clusters)
      mtcars$cluster <- as.factor(clusters$cluster)
      analysis_data(table(mtcars$cluster))
    }
  })
  
  output$analysis_results <- renderPrint({
    analysis_data()
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```


